---
title: "615HW5"
output: pdf_document
date: "2024-10-05"
author: "Chang Lu"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)  
library(kableExtra)
library(tidyverse)
library(rstanarm)
```

## 615 Strawberry data cleaning
```{r}
#| label: read data - glimpse 

# Load the strawberry dataset and take a glimpse at the data
strawberry_data <- read_csv("strawberries25_v3.csv", col_names = TRUE)
glimpse(strawberry_data)
```

### Explanation
Dataset Overview: The `strawberry_data` contains 12,699 rows and 21 columns. Columns include information like date, location, values, and coefficients of variation.

### Remove Columns with Single Value in All Rows

```{r}
#|label: function def - drop 1-item columns

# Function to remove columns with only a single unique value
drop_one_value_col <- function(df) {   
  drop <- NULL
  
  # Test each column for a single value
  for (i in 1:ncol(df)) {    
    if (n_distinct(df[[i]]) == 1) {
      drop <- c(drop, i)
    }
  }
  
  # Report the result - names of columns dropped
  if (is.null(drop)) {
    return(df)
  } else {
    print("Columns dropped:")
    print(colnames(df)[drop])
    df <- df[, -drop]
  }
}

# Use the function to remove single-value columns
strawberry_data <- drop_one_value_col(strawberry_data)
glimpse(strawberry_data)
```

### Separate Composite Columns

```{r}
#|label: split Data Item

# Split 'Data Item' column into two parts using delimiter '-'
strawberry_data <- strawberry_data |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = "-",
                         names = c("column1", "column2"),
                         too_many = "merge",
                         too_few = "align_start")

# Further split 'column1' into 'Fruit' and 'Category' using delimiter ','
strawberry_data <- strawberry_data |>
  separate_wider_delim(  cols = `column1`,
                         delim = ",",
                         names = c("Fruit", "Category"),
                         too_many = "merge",
                         too_few = "align_start")

# Trim white spaces from the split columns
strawberry_data$Fruit <- str_trim(strawberry_data$Fruit, side = "both")
strawberry_data$Category <- str_trim(strawberry_data$Category, side = "both")
strawberry_data$column2 <- str_trim(strawberry_data$column2, side = "both")

# Remove columns with single-value again after modifications
strawberry_data <- drop_one_value_col(strawberry_data)
unique(strawberry_data$Category)
```

### Clean Data in the 'Category' Column

```{r}
#|label: Clean data in the Category

# Extract specific attributes from 'Category' and assign them to new columns
strawberry_data <- strawberry_data %>%
  mutate(Marketing_channels = ifelse(str_detect(Category, "FRESH MARKET|PROCESSING"),
                                      str_extract(Category, "FRESH MARKET|PROCESSING"), NA)) %>%
  mutate(Utilizations = ifelse(str_detect(Category, "UTILIZED"),
                               str_extract(Category, "UTILIZED"), NA)) %>%
  mutate(Method = ifelse(str_detect(Category, "ORGANIC"),
                         str_extract(Category, "ORGANIC"), NA)) %>%
  mutate(Class = ifelse(str_detect(Category, "BEARING"),
                        str_extract(Category, "BEARING"), NA)) %>%
  mutate(Measurement = ifelse(str_detect(Category, "NOT SOLD"),
                              str_extract(Category, "NOT SOLD"), NA)) %>% 
  select(1:match("County ANSI", names(.)), Class, Method, Marketing_channels, Utilizations, Measurement, column2, everything()) %>% 
  select(-Category)
```

### Clean Data in 'column2'

```{r}
#|label: Clean column2

# Split 'column2' into 'Measurement1' and 'Metric1'
strawberry_data <- strawberry_data %>%
  mutate(
    Measurement1 = ifelse(str_detect(column2, ","),
                          str_split_fixed(column2, ",", 2)[, 1],
                          column2),  
    Metric1 = ifelse(str_detect(column2, ","),
                    str_split_fixed(column2, ",", 2)[, 2],
                    NA) 
  ) %>% 
  select(1:match("Measurement", names(.)), Measurement1, Metric1, everything()) %>% 
  select(-column2)
```

### Classify Strings in Their Positions

```{r}
#|label: classify strings

# Extract specific strings from 'Metric1' to 'Metric' and 'Remark'
strawberry_data <- strawberry_data %>%
  mutate(Metric = str_extract(Metric1, "(?<=,|^)[^,]*MEASURED IN[^,]*(?=,|$)")) %>%
  mutate(Remark = str_remove_all(Metric1, "(?<=,|^)[^,]*MEASURED IN[^,]*(,)?")) %>%
  mutate(Remark = str_trim(str_replace_all(Remark, "^,|,$|,,", ""))) %>%
  select(1:match("Measurement1", names(.)), Metric, Remark, everything())  %>% 
  select(-Metric1)

# Create 'Category' column by combining 'Measurement' and 'Measurement1'
strawberry_data <- strawberry_data %>%
  mutate(Category = ifelse(is.na(Measurement), 
                           Measurement1,
                           paste(Measurement, Measurement1, sep = " "))) %>% 
  select(1:match("Utilizations", names(.)), Category, everything()) %>% 
  select(-Measurement, -Measurement1)

# Trim whitespace
strawberry_data$Metric <- str_trim(strawberry_data$Metric, side = "both")
strawberry_data$Category <- str_trim(strawberry_data$Category, side = "both")

# Replace empty strings in 'Remark' with NA
strawberry_data <- strawberry_data %>% mutate(across(Remark, ~ na_if(., "")))
glimpse(strawberry_data)
```

### Separate Domain and Domain Category

```{r}
#|label: separate domain category

# Clean up 'Domain Category' by removing redundant text and characters
strawberry_data <- strawberry_data %>%
  mutate(`Domain Category` = ifelse(str_detect(`Domain Category`, ":"),
                              str_split_fixed(`Domain Category`, ":", 2)[, 2],
                              `Domain Category`)) %>%
  mutate(`Domain Category` = str_replace_all(`Domain Category`, "[\\(\\)]", ""))

strawberry_data$`Domain Category` <- str_trim(strawberry_data$`Domain Category`, side = "both")

# Separate 'Domain Category' into specific chemical and numbers
strawberry_data <- strawberry_data %>%
  mutate(Chemical_Number = ifelse(str_detect(`Domain Category`, "="),
                                  str_split_fixed(`Domain Category`, "=", 2)[, 2], 
                                  NA)) %>%
  mutate(`Domain Category` = ifelse(str_detect(`Domain Category`, "="),
                              str_split_fixed(`Domain Category`, "=", 2)[, 1],  
                              `Domain Category`)) %>% 
  select(1:match("Domain Category", names(.)), Chemical_Number, everything())

# Trim whitespace
strawberry_data$`Domain Category` <- str_trim(strawberry_data$`Domain Category`, side = "both")
strawberry_data$Chemical_Number <- str_trim(strawberry_data$Chemical_Number, side = "both")

# Remove 'CHEMICAL' from the 'Domain' column
strawberry_data <- strawberry_data %>%
  mutate(Domain = str_replace(Domain, "CHEMICAL, ", ""))
glimpse(strawberry_data)
```

### Transfer Data Types to Correct Types

```{r}
#|label: transfer data types

# Replace specific strings in 'Value' and 'CV (%)' columns and convert to numeric
strawberry_data <- strawberry_data %>%
  mutate(Value = ifelse(Value %in% c("(D)", "(NA)"), NA, Value)) %>%
  mutate(Value = ifelse(Value == "(Z)", "0.0005", Value)) %>% 
  mutate(Value = str_replace_all(Value,",","")) %>% 
  mutate(Value = as.numeric(Value))

strawberry_data <- strawberry_data %>%
  mutate(`CV (%)` = ifelse(`CV (%)` %in% c("(D)", "(NA)"), NA, `CV (%)`)) %>%
  mutate(`CV (%)` = ifelse(`CV (%)` == "(L)", "0.05", `CV (%)`)) %>% 
  mutate(`CV (%)` = ifelse(`CV (%)` == "(H)", "99.95", `CV (%)`)) %>% 
  mutate(`CV (%)` = str_replace_all(`CV (%)`,",","")) %>% 
  mutate(`CV (%)` = as.numeric(`CV (%)`))
glimpse(strawberry_data)
```

### Separate the Data into Categories

```{r}
#|label: separate data categories

# Split the table into 'census' and 'survey'
strawberry_census <- strawberry_data %>% filter(Program == "CENSUS")
strawberry_survey <- strawberry_data %>% filter(Program == "SURVEY")
strawberry_census <- drop_one_value_col(strawberry_census)
strawberry_survey <- drop_one_value_col(strawberry_survey)

# Split the 'census' table into 'organic' and 'non-organic'
strawberry_organic <- strawberry_census %>% filter(Domain == "ORGANIC STATUS")
strawberry_organic <- drop_one_value_col(strawberry_organic)
strawberry_non_organic <- strawberry_census %>% filter(!Domain == "ORGANIC STATUS")
strawberry_non_organic <- drop_one_value_col(strawberry_non_organic)

# Split the 'survey' table into 'chemical' and 'non-chemical'
strawberry_chemical <- strawberry_survey %>% filter(!Domain == "TOTAL")
strawberry_chemical <- drop_one_value_col(strawberry_chemical)
strawberry_non_chemical <- strawberry_survey %>% filter(Domain == "TOTAL")
strawberry_non_chemical <- drop_one_value_col(strawberry_non_chemical)
```

### Estimate the NA in Value and CV(%)

```{r}
#|label: estimate missing values

# Function to estimate missing values in a given dataset using linear regression
estimate_na_values <- function(df, value_model_formula, cv_model_formula = NULL) {
  df <- df %>% mutate(original_order = row_number())
  # Value estimation
  with_value <- df %>% filter(!is.na(Value))
  missing_value <- df %>% filter(is.na(Value))
  value_model <- lm(value_model_formula, data = with_value)
  missing_value <- missing_value %>%
    mutate(Value = round(exp(predict(value_model, newdata = missing_value)), 1))
  df_filled <- bind_rows(with_value, missing_value) %>% arrange(original_order)
  # CV estimation
  if (!is.null(cv_model_formula)) {
    with_cv <- df_filled %>% filter(!is.na(`CV (%)`))
    missing_cv <- df_filled %>% filter(is.na(`CV (%)`))
    cv_model <- lm(cv_model_formula, data = with_cv)
    missing_cv <- missing_cv %>%
      mutate(`CV (%)` = round(predict(cv_model, newdata = missing_cv), 2))
    df_filled <- bind_rows(with_cv, missing_cv) %>% arrange(original_order)
  }
  return(df_filled %>% select(-original_order) %>% mutate(Value = round(Value, 1)))
}

# Estimate missing values for 'strawberry_organic'
strawberry_organic <- estimate_na_values(strawberry_organic, log(Value) ~ factor(Year) + State + Category)

# Estimate missing values for 'strawberry_non_organic'
strawberry_non_organic <- estimate_na_values(strawberry_non_organic, Value ~ State + Category + Domain + `Domain Category`, `CV (%)` ~ State + Category + Domain + `Domain Category`)

# Estimate missing values for 'strawberry_chemical'
strawberry_chemical <- estimate_na_values(strawberry_chemical, log(Value) ~ factor(Year) + State + Category + Domain)

# Estimate missing values for 'strawberry_non_chemical'
strawberry_non_chemical <- estimate_na_values(strawberry_non_chemical, log(Value + 1) ~ factor(Year) + State + Category + Period)
```

### Write Cleaned Data to CSV Files

```{r}
#|label: write csv

# Write cleaned datasets to new CSV files
write.csv(strawberry_organic, "strawberry_organic.csv", row.names = FALSE)
write.csv(strawberry_non_organic, "strawberry_non_organic.csv", row.names = FALSE)
write.csv(strawberry_chemical, "strawberry_chemical.csv", row.names = FALSE)
write.csv(strawberry_non_chemical, "strawberry_non_chemical.csv", row.names = FALSE)
```
